---
title: "exercises"
author: "luke reding"
date: "2/8/2017"
output: html_document
---

### Question 1

Load the `tidyverse` package.

```{r}



```


### Question 2

Use the `read_csv` function to read in the common garden video spreadsheet. Save it to an object called `common_garden`. Check to to make sure all the data types are correct.

_Hint:_ If you don't know where the spreadsheet is, use `file.choose()` to find it.

```{r}



```



### Question 3

Do a web search for how to find (a) the number of rows and (b) the number of columns in your `common_garden` dataframe. Enter in the R code that returns the number of rows and columns below.


```{r}



```


### Question 4

Use `ggplot` to plot the tank condition on the x-axis and the total number of male-male chases on the y axis.

_Hint_: Remember the basics of using ggplot: To set up a ggplot, you need a `data` argument and a `mapping` argument. The `mapping` argument always looks like `mapping = aes(x = some_column_name, y = some_other_column_name)`. Then type `+` followed by whatever `geoms` you want to use, for example, `geom_point()`. 

Follow the code from the lecture if you get confused, substituting the arguments.

```{r}



```



### Question 5

Plot the relationship between male-male chases (x-axis) and total male courting (y-axis) as a scatterplot.

Again, follow the example scatterplot from the lecture if you're confused.

```{r}



```


### Question 6

Perform a correlation test to test whether there is statistically significant relationship between total male-male chases and total male courtships.

```{r}


```


### Question 7

boxplot of foraging in males vs females




female-female chases by treatment









